A Formal Interface for Concurrent Search Structure Templates

### 2024-11-14 concurrency/template paper/paper.tex
outlining paper on formal templates

%\usepackage{txfonts}
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%% http://ctan.org/pkg/subcaption
\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{tikzit}
\input{box.tikzstyles}
\usetikzlibrary{fit}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{parcolumns}
\usepackage{scalerel}
%\usepackage{lstlangcoq}
\usepackage[edges]{forest}
\renewcommand{\lstlistingname}{Figure}

@@

%https://tex.stackexchange.com/questions/112634/how-to-draw-a-diagram-with-boxes-of-text-inside-each-other-and-have-them-laid-ou
\newcommand\TBox[2][]{%
  \tikz\node[draw,align=left,#1] {#2};\hskip2pt}

@@

%%% The following is specific to CPP '24 and the paper
%%% 'Compositional Verification of Concurrent C Programs with Search Structure Templates'
%%% by Duc-Than Nguyen, Lennart Beringer, William Mansky, and Shengyi Wang.
%%%
\setcopyright{acmlicensed}
\acmDOI{10.1145/3636501.3636940}
\acmYear{2024}
\copyrightyear{2024}
\acmSubmissionID{poplws24cppmain-p2-p}
\acmISBN{979-8-4007-0488-8/24/01}
\acmConference[CPP '24]{Proceedings of the 13th ACM SIGPLAN International Conference on Certified Programs and Proofs}{January 15--16, 2024}{London, UK}
\acmBooktitle{Proceedings of the 13th ACM SIGPLAN International Conference on Certified Programs and Proofs (CPP '24), January 15--16, 2024, London, UK}
\received{2023-09-19}
\received[accepted]{2023-11-25}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Compositional Verification of Concurrent C Programs with Search Structure Templates}

\author{Duc-Than Nguyen}
\orcid{0000-0002-6810-897X}
\affiliation{%
	\institution{University of Illinois at Chicago}
	\city{}
	\country{USA}
}
\email{dnguye96@uic.edu}

\author{Lennart Beringer}
\orcid{0000-0002-1570-3492}
\affiliation{%
	\institution{Princeton University}
	\city{}
	\country{USA}
}
\email{eberinge@cs.princeton.edu}

\author{William Mansky}
\orcid{0000-0002-5351-895X}
\affiliation{%
	\institution{University of Illinois at Chicago}
	\city{}
	\country{USA}
}
\email{mansky1@uic.edu}

\author{Shengyi Wang}
\orcid{0000-0002-2286-8703}
\affiliation{%
	\institution{Princeton University}
	\city{}
	\country{USA}
}
\email{shengyiw@princeton.edu}
%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort. In this paper, we implement the template approach in the Verified Software Toolchain (VST), and use it to prove correctness of C implementations of fine-grained concurrent data structures. This involves translating code, specifications, and proofs to the idiom of C and VST, and gives us another look at the requirements and limitations of the template approach.
We encounter several questions about the boundaries between template and data structure, as well as some common data structure operations that cannot naturally be decomposed into templates. Nonetheless, the approach appears promising for modular verification of real-world concurrent data structures.
\end{abstract}

@@

\section{Introduction}
\label{sec:introduction}
Concurrent search structure templates~\cite{shasha1988concurrent} are a technique for describing and implementing concurrency patterns (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) that can then be specialized to various search data structures (e.g., linked list, hashtable, B-tree) compositionally. Krishna et al. ~\cite{templates} repurposed templates as a compositional \emph{verification} technique, separating the proof of a concurrent access method from the proof of the underlying data structure. The concurrency templates are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency. In theory, this could allow us to prove the correctness of $n$ concurrency patterns and $m$ (single-threaded) data structure implementations, and immediately obtain $n \times m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. %\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}

@@

Our specific contributions are:
\begin{itemize}
	\item We reimplement the template approach independently of flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/substructure''.
	\item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.
	\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
	\item We give a precise description of search structure templates, and identify places where it is difficult in practice to preserve the boundary between template and data structure: for instance, both node creation and rotation in binary search trees operate on concurrent and data-structure-specific aspects in ways that do not cleanly decompose.
\end{itemize}

\subsection*{Related Work}
Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}. The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach. Some of the more complex data structures verified include Java's \lstinline{ConcurrentSkipListMap}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}. Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeting an implementation language (C) rather than a toy language or algorithmic representation. For instance, Iris's RustBelt instance~\cite{rustbelt} targets a core calculus based on Rust's intermediate language and has been used to verify several interesting concurrent programs, but does not provide guarantees about actual running Rust code.

@@

It is also worth mentioning recent work that extends the reach of template-style reasoning. Later work on search structure templates by Patel et al.~\cite{template-multi} applies the template approach to multicopy search structures, where there may be more than one node containing the target key. Feldman et al.~\cite{feldman2020proving} take an approach similar to templates to verify search structures with highly optimistic concurrency patterns, where the data structure may be restructured by other threads during traversal. Their technique could potentially be applied to prove correctness of much more complex \lstinline{traverse} operations than those we describe.

\section{Background}
\subsection{Concurrent Search Structure Templates}

@@

\subsubsection{Logical atomicity} 
\label{logatom}
%In this section, we show how to verify the templates presented above, by proving that each \lstinline{traverse} function meets its specification. %The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}.

@@

To apply the template approach to real-world code, we instead use the Verified Software Toolchain (VST)~\cite{plcc}, a separation-logic-based verifier for C programs. VST is also built in Coq and is connected to the CompCert verified C compiler~\cite{compcert}, allowing it to guarantee that proved properties will hold on compiled code. Recent work on VST~\cite{iris-vst-arxiv} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications. In this paper, we use these extensions to reconstruct the template approach in VST and apply it to real C programs. %clean this up a little

\section{Search Structure Templates in VST}

@@

A search structure template is an implementation of these three functions, written in a particularly restricted way outlined in Figure~\ref{templates-diagram}. Aside from some top-level management code (e.g., acquiring and releasing an initial lock), all that each function should do is call two functions: a helper function called \lstinline{traverse}, which is specific to the template and implements its synchronization mechanism, and a concurrency-unaware function (e.g. \lstinline{insertOp}) provided by the target data structure. (In fact, Krishna et al. implemented all three operations as a single function \lstinline{decisiveOp}, parameterized by a function that takes the target operation type as input and performs the appropriate data structure operation; for ease of adaptation to C, we write three different functions, but their structures are identical.) A template consists of an implementation of the top-level operations and the \lstinline{traverse} function used to define them; a data structure consists of an implementation of \lstinline{insertOp}, etc., as well as a function \lstinline{findNext} that is used to implement \lstinline{traverse}. The interface between them is a set of standard specifications for \lstinline{insertOp}, etc., and \lstinline{findNext}: if a template's functions can be verified assuming these standard specifications, and a data structure's implementations can be shown to meet those specifications, then the template and the data structure can be combined to yield a verified concurrent search structure.

@@

\subsubsection{Verifying the Lock-Coupling Template}
\label{traverse_proof_lock}

@@

\subsubsection{Verifying the Give-Up Template}
\label{traverse_proof_giveup}
The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{ \color{specblue}
		\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\texttt{pn} \mapsto (p, n) \ast  
				\inFP(n)   
			\end{array}  
			\Big \vert \ \treerep\ (m)
		}  
		{\rangle}
	}
	\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{ \color{specblue}
		\scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\mathit{res}.\ \exists \  n', v, \mathit{range}. \\ 
				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast \texttt{k} \in \mathit{range} \ \ast 
				\\ 
				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ 
				\\ 
				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathit{range})
			\end{array}  
			\Bigg \vert \ \treerep\ (m)
		}  
		{\rangle}
	}
\end{mathpar}

@@

\section{Verifying Binary Search Trees using Templates}
\label{BST_proof}
In this section, we demonstrate how to instantiate the templates with specific data structures, giving us verified C implementations of concurrent data structures. 
As we saw in the previous section, the top-level specifications of data structure operations have already been verified at the template level; all we need to do now is provide implementations of data-structure-specific functions (e.g. \lstinline{findNext}) that satisfy the right specifications to instantiate the template proofs. Once we have chosen a template and a data structure, we should not need to do any further reasoning about specific concurrency patterns.

Our target data structure is a binary search tree (BST) implemented in C. A lock-coupling BST was already written verified as a demonstration of VST's concurrency capabilities~\cite{bst-conc}. We refactored it into \lstinline{findNext}, \lstinline{traverse}, \lstinline{insertOp}, etc. as required by the template approach, and then re-verified it using the template; we then replaced \lstinline{traverse} with the give-up version and verified the resulting give-up BST as well. We were able to obtain verified \lstinline{insert} and \lstinline{lookup} operations using this technique; for \texttt{delete}, we found a mismatch between the template approach and BST deletion, which we discuss in Section \ref{internal_reorganize}.

\subsection{Instantiating \lstinline{findNext} and \lstinline{insertOp}}
We instantiate the templates with the binary search tree by defining and verifying the helper functions \texttt{findNext} (used in \lstinline{traverse}) and \texttt{insertOp} (used in \texttt{insert}). As expected, their implementations are strictly sequential, their specifications are ordinary non-atomic Hoare triples (see Figures \ref{fig:findNext_lock} and \ref{fig:insertOp}), and their proofs do not require any concurrency reasoning.
\begin{figure*}[!ht]
	\centering
	\begin{mathpar}
		{\color{specblue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)
			\end{array}
			\right\}
		} \vspace{-0.85em} 
		\\ 
		\texttt{findNext(pn, k)} 
		\vspace{-0.85em} 
		\\
		{\color{specblue}
			\left\{\mathit{res}. \ \exists \  n'.
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)  \ \ast \\\  
				\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k')
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)}
	\label{fig:findNext_lock}
\end{figure*}

\begin{figure*}[h]
	\begin{subfigure}{\textwidth}
		\centering
		\begin{mathpar}
			{\color{specblue}
				\left\{ 
				\begin{array}{c}
					\ \texttt{pn->n} \mapsto \texttt{NULL}  
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\ 
			\texttt{insertOp(pn, k, v)} 
			\vspace{-0.85em} 
			\\
			{\color{specblue}
				\left\{\exists \  l\ r\ \mathit{lk}_1\ \mathit{lk}_2.
				\begin{array}{l}
					\ \texttt{pn->n->t} \mapsto (\texttt{k}, \texttt{v}, l, r) \ \ast \ \\ 
					\ l \mapsto (\texttt{NULL}, \mathit{lk}_1)  \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2)
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the lock-coupling template}
		\label{fig:insertOp_lock}
	\end{subfigure}
	%insertOp for give-up of give-up
	\begin{subfigure}{\textwidth}
		\begin{mathpar}
			{\color{specblue}
				\left\{ 
				\begin{array}{c}
					\texttt{pn->n}\mapsto \texttt{NULL}  \ast  \texttt{pn->n->min}\mapsto n_1  \ast  \texttt{pn->n->max}\mapsto n_2
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\
			\texttt{insertOp(pn, k, v)}
			\vspace{-0.85em} 
			\\
			{\color{specblue}
				\left\{\exists \ l\ r\ \mathit{lk}_1\ \mathit{lk}_2. 
				\begin{array}{c}
					\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r)  \ast  \texttt{pn->n->min}\mapsto n_1 \ast  \texttt{pn->n->max}\mapsto n_2  \ \ast 
					\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (n_1, \texttt{k})) \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, n_2))
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the give-up template}
		\label{fig:insertOp_giveup}
	\end{subfigure}
	\caption{Specification of \texttt{insertOp} for the lock-coupling and give-up templates}
	\label{fig:insertOp}
\end{figure*}

The contents of a BST node are defined in C as:
\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
	typedef struct node {
		int key; 
		void *value; 
		struct node_t *left, *right;
	} node;
\end{lstlisting}
The BST's \lstinline{findNext} function (Figure \ref{findnext}) indicates whether the target key is in the current node, its left subtree, or its right subtree. The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with key $k'$, value $v'$, and pointers to the left $l$ and right $r$ child nodes. The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the node's key $k'$, and the next node $n'$ is the left child $l$; \lstinline{k} is greater than $k'$ and $n'$ is the right child $r$; or \lstinline{k} is exactly $k'$ and $n'$ is the original node $n$, in which case the return value is 0. The proof is straightforward.
\begin{figure}[ht]
	\begin{minipage}{0.45\textwidth}
	\lstinputlisting[language=C, style=myStyle]{findnext.c} 
	\end{minipage}
	\caption{C implementation of \lstinline{findNext} for the binary search tree (all templates)}
	\label{findnext}
\end{figure}

The BST's \lstinline{insertOp} function allocates a new \lstinline{node} struct at an empty leaf, as well as two new empty leaves to serve as its children. The \texttt{insertOp} specification (Figure \ref{fig:insertOp_lock}) takes as input a node whose \lstinline{t} field is \lstinline{NULL}, i.e., an empty leaf node. It inserts a new node with the specified key \texttt{k} and value \texttt{v} at that leaf, complete with two new empty child nodes with corresponding locks, $(\texttt{NULL}, \mathit{lk}_1)$ and $(\texttt{NULL}, \mathit{lk}_2)$, which are pointed to by the $l$ and $r$ of the current node.

Unfortunately, while this specification suffices for the lock-coupling template, it cannot be used as is for the give-up template: because we allocate new empty leaf nodes, we must allocate \lstinline{node_t} structures, whose implementation in the give-up template has extra fields that do not appear in the lock-coupling template. As shown in Figure \ref{fig:insertOp_giveup}, a give-up template node (even an empty one!) has additional \lstinline{min} and \lstinline{max} fields, and these fields are needed to compute the bounds of the new child nodes. In a BST, if the current node's range is (\lstinline{min, max}) and its key is \lstinline{k}, its left child will have range (\texttt{min, k}) and its right child will have range (\texttt{k, max}). It may be possible to remedy this discrepancy by adding an ``allocate empty node'' function to the templates, which is then called by \lstinline{insertOp}, but this would still complicate the approach: originally, we expected that templates would be parameterized by data structure implementations but not vice versa. However, this two-way dependency seems fundamental when the layout of the node structure depends on the concurrency control mechanism we choose. For now, we break the abstraction boundary and verify two different versions of \lstinline{insertOp}, one for the lock-coupling template and one for the give-up template, which reduces the proof-reuse benefits of the template approach.

%The \texttt{insert} specification for the give-up template is identical to that of the lock-coupling template, up to changes in the definitions of $\nodeboxrep\ \texttt{p}$ and $\treerep \ \texttt{m}$. %Essentially, the \texttt{insert} function's specification comprises two main components: local and public preconditions and postconditions, as previously mentioned. Moreover, we must demonstrate that it is possible to atomically update the tree's abstract state from an unknown \texttt{m} to $\texttt{m[k} \mapsto \texttt{v])}$.
%The local preconditions and postconditions, containing $\nodeboxrep \ \texttt{p}$, depict the binary search tree's concrete state, where the root node $\texttt{p}$ points to the node's lock and footprint predicate, signifying that the root node is located within the search structure's footprint. In contrast, the public preconditions and postconditions are denoted by the predicate  $\treerep \ \texttt{m}$, which is identical to the public predicates of the \texttt{traverse} function in the give-up template.

%The proof for the \texttt{insert} operation begins with the preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$. We assign \texttt{pn->n} to point to the same memory address as \texttt{*p} (line 3 in Figure \ref{insert_giveup}). In Section \ref{traverse_proof_giveup}, we discuss providing the identical preconditions for the \texttt{traverse} function, which encompasses the local precondition $\infp\texttt{(pn->n)}$ and the public precondition $\treerep\ \texttt{m}$. Subsequently, we apply the \texttt{traverse} specification from Section \ref{traverse_proof_giveup}.

%Analogous to the lock-coupling template's \texttt{insert} operation, there are two possible branches representing the outcome following the execution of \texttt{traverse}. Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ \texttt{(m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.

\subsection{Templates and Internal Reorganization}
\label{internal_reorganize}
The \lstinline{delete} operation on binary search trees highlights a limitation of the template approach as we have described it. The \lstinline{insert} and \lstinline{lookup} operations can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insertOp} or \lstinline{lookupOp}). In most of the prior template examples, deletion can be decomposed similarly, by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation. When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure or code?
rotating three-node sections of the tree until the node to be deleted is at a leaf. This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them).

>>>>

%\usepackage{txfonts}
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%% http://ctan.org/pkg/subcaption
\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{tikzit}
\input{box.tikzstyles}
\usetikzlibrary{fit}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{parcolumns}
\usepackage{scalerel}
%\usepackage{lstlangcoq}
\usepackage[edges]{forest}
\usepackage{balance}
\renewcommand{\lstlistingname}{Figure}

@@

@@

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{A Formal Interface for Concurrent Search Structure Templates}

\author{Duc-Than Nguyen}
\orcid{0000-0002-6810-897X}
\affiliation{%
	\institution{University of Illinois Chicago}
	\city{}
	\country{USA}
}
\email{dnguye96@uic.edu}

\author{William Mansky}
\orcid{0000-0002-5351-895X}
\affiliation{%
	\institution{University of Illinois Chicago}
	\city{}
	\country{USA}
}
\email{mansky1@uic.edu}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}

\end{abstract}

@@

\section{Introduction}
\label{sec:introduction}
Search structure templates~\cite{templates} were introduced by Krishna et al. as an approach to modular verification of concurrent data structures. The approach consists of a programming pattern and specification style that decompose concurrent data structures into \emph{concurrency templates} (lock coupling, give-up) and \emph{sequential data structures} (linked list, hash table). Later work in the same line extends the template approach to adding multicopy support (e.g., log-structure merging)~\cite{}, or combining thread-safe node implementations into larger data structures~\cite{}. However, these applications are qualitatively different from the original promise of the approach: that we can decompose both the implementation and the proof of a concurrent data structure into a \emph{concurrency-unaware} data structure component and a \emph{data-structure-unaware} concurrency component. This form of modularity is, to the best of our knowledge, unique in the literature, at least when the concurrent component is fine-grained (i.e., implemented with a lock per node or lock-free operations rather than one big lock). It also poses unique challenges in both implementation and specification/verification.

However, while \citet{templates} lay out the high-level ideas of the template approach and work through several examples, they do not formally define the approach. In particular, a modularization technique should have a well-defined \emph{interface} for each of its components, so that we know that components written and verified according to the technique can be successfully combined. In theory, given $m$ data structure components and $n$ concurrency templates, we should immediately be able to obtain $m \times n$ verified concurrent data structures. The original paper by Krishna et al. does not realize this (nor does it claim to): it presents three verified templates and five verified data structures, but each data structure is combined with only one template. In fact, close examination of the examples shows that different templates use different and incompatible specifications for the data structure component. In this paper, we rectify this by presenting a \textbf{formal interface for concurrent search structure templates}, implemented as a library in C and a collection of typeclasses in Coq, that guarantees $m \times n$ verified data structures from $m + n$ components. More specifically:
\begin{itemize}
\item We describe an approach to \textbf{implementing} data structures and synchronization mechanisms so that they can be freely combined. This is a surprisingly challenging task: ideally, the data structure implementations should not explicitly leave space for concurrency metadata like locks, and the concurrency mechanism should not come between pieces of the data structure. We discuss several undesirable approaches and their relationships to object-orientation patterns, and show that the most desirable approach corresponds to \emph{traits} or \emph{aspects}, though it can also be implemented in C.
\item We give a formal \textbf{interface} for data structure components and concurrency templates, in the form of separation logic specifications for each of the relevant functions. These specifications follow the style of \citet{templates}, but we clarify which specifications should be considered the official interface and which are derived or implementation-specific, as well as ensuring that data structure components are never assumed to include concurrent features and concurrency templates never rely on data structure details. We verify the top-level concurrent data structure operations (\lstinline{insert} and \lstinline{lookup}) against these specifications, guaranteeing that they work correctly for any data structure component and template that satisfy the interface.
\item We exhibit two data structures (linked list and BST) and two concurrency templates (lock coupling and give-up) that satisfy our interface, and from them \textbf{freely obtain four verified concurrent data structures}, implemented in C and verified in VST. %The BST is particularly interesting because...
\end{itemize}

\ignore{
Concurrent search structure templates~\cite{shasha1988concurrent} are a technique for describing and implementing concurrency patterns (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) that can then be specialized to various search data structures (e.g., linked list, hashtable, B-tree) compositionally. Krishna et al. ~\cite{templates} repurposed templates as a compositional \emph{verification} technique, separating the proof of a concurrent access method from the proof of the underlying data structure. The concurrency templates are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency. In theory, this could allow us to prove the correctness of $n$ concurrency patterns and $m$ (single-threaded) data structure implementations, and immediately obtain $n \times m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. %\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}

@@

Our specific contributions are:
\begin{itemize}
	\item We reimplement the template approach independently of flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/substructure''.
	\item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.
	\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
	\item We give a precise description of search structure templates, and identify places where it is difficult in practice to preserve the boundary between template and data structure: for instance, both node creation and rotation in binary search trees operate on concurrent and data-structure-specific aspects in ways that do not cleanly decompose.
\end{itemize}
}

\ignore{
\paragraph{Related Work}
Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}. The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach. Some of the more complex data structures verified include Java's \lstinline{ConcurrentSkipListMap}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}. Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeting an implementation language (C) rather than a toy language or algorithmic representation. For instance, Iris's RustBelt instance~\cite{rustbelt} targets a core calculus based on Rust's intermediate language and has been used to verify several interesting concurrent programs, but does not provide guarantees about actual running Rust code.

@@

It is also worth mentioning recent work that extends the reach of template-style reasoning. Later work on search structure templates by Patel et al.~\cite{template-multi} applies the template approach to multicopy search structures, where there may be more than one node containing the target key. Feldman et al.~\cite{feldman2020proving} take an approach similar to templates to verify search structures with highly optimistic concurrency patterns, where the data structure may be restructured by other threads during traversal. Their technique could potentially be applied to prove correctness of much more complex \lstinline{traverse} operations than those we describe.}

\ignore{
\section{Background}
\subsection{Concurrent Search Structure Templates}

@@

\paragraph{Logical atomicity} 
\label{logatom}
%In this section, we show how to verify the templates presented above, by proving that each \lstinline{traverse} function meets its specification. %The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}.

@@

To apply the template approach to real-world code, we instead use the Verified Software Toolchain (VST)~\cite{plcc}, a separation-logic-based verifier for C programs. VST is also built in Coq and is connected to the CompCert verified C compiler~\cite{compcert}, allowing it to guarantee that proved properties will hold on compiled code. Recent work on VST~\cite{iris-vst-arxiv} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications. In this paper, we use these extensions to reconstruct the template approach in VST and apply it to real C programs. %clean this up a little
}

\section{Coding Search Structure Templates}
2-by-2 grid of concurrent node implementations (bst+lock, bst+lock+range, list+lock, list+lock+range)

which pieces are data structure and which are template? (paper assumes locks and ranges are provided by data structure)

difficulty of actually implementing things this way -- even in C++, we'd have to write subclasses for BST-coupling-node, BST-giveup-node, LL-coupling-node, LL-giveup-node, and in Java those classes would have to have code in them

sketch the four approaches, describe why only one works for us?

some of the tricky pieces to separate: node creation/allocation of initial metadata, updating metadata on insert, ...

highlight the interface: which functions must a data structure provide? which functions must a template provide? which functions are written once generically, parameterized by both data structure and template?

\section{Specifying the Template Interface}
standard Hoare triples for each data structure function

atomic triples for each template function, with data structure instance as parameter

top-level specs proved once-and-for-all, with data structure and template instance as parameters

\section{Data Structure Instances: Linked List and Binary Search Tree}

\section{Template Instances: Give-Up and Lock Coupling}

\section{Evaluation}

\ignore{\section{Search Structure Templates in VST}

@@

A search structure template is an implementation of these three functions, written in a restricted way outlined in Figure~\ref{templates-diagram}. Aside from some top-level management code (e.g., acquiring and releasing an initial lock), all that each function should do is call two functions: a helper function called \lstinline{traverse}, which is specific to the template and implements its synchronization mechanism, and a concurrency-unaware function (e.g. \lstinline{insertOp}) provided by the target data structure. (In fact, Krishna et al. implemented all three operations as a single function \lstinline{decisiveOp}, parameterized by a function that takes the target operation type as input and performs the appropriate data structure operation; for ease of adaptation to C, we write three different functions, but their structures are identical.) A template consists of an implementation of the top-level operations and the \lstinline{traverse} function used to define them; a data structure consists of an implementation of \lstinline{insertOp}, etc., as well as a function \lstinline{findNext} that is used to implement \lstinline{traverse}. The interface between them is a set of standard specifications for \lstinline{insertOp}, etc., and \lstinline{findNext}: if a template's functions can be verified assuming these standard specifications, and a data structure's implementations can be shown to meet those specifications, then the template and the data structure can be combined to yield a verified concurrent search structure.

@@

\paragraph{Verifying the Lock-Coupling Template}
\label{traverse_proof_lock}

@@

\paragraph{Verifying the Give-Up Template}
\label{traverse_proof_giveup}
The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{ \color{specblue}
		\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\texttt{pn} \mapsto (p, n) \ast  
				\inFP(n)   
			\end{array}  
			\Big \vert \ \treerep\ (m)
		}  
		{\rangle}
	}
	\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{ \color{specblue}
		\scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\mathit{res}.\ \exists \  n', v, \mathit{range}. \\ 
				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast \texttt{k} \in \mathit{range} \ \ast 
				\\ 
				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ 
				\\ 
				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathit{range})
			\end{array}  
			\Bigg \vert \ \treerep\ (m)
		}  
		{\rangle}
	}
\end{mathpar}

@@

}

\section{Conclusion and Future Work}
The framework we have presented fulfills the true promise of the concurrency template approach: each data structure instance is implemented and verified without any awareness of concurrency or metadata, each template instance is implemented and proved without any knowledge of the implementation of the data structure, and we obtain verified concurrent data structures immediately by instantiating our top-level theorems with any combination of data structure and template instances. We accomplished this by careful attention to which components belong to the data structure and which to the template, and by working to disentangle the two in both the C implementation and the Coq specification and proof. Our interfaces are general enough that it should be easy to add more data structures and templates, including concurrency patterns that do not depend on locks (e.g., optimistic concurrency control with fine-grained atomics). We have seen that the concurrency template must be implemented in a very specific way, with metadata stored separately from the data structure itself, in order to sufficiently separate concerns; in future work, we are interested to see whether languages with more flexible composition mechanisms, like traits in Rust and Scala, lend themselves to more natural implementations of this composition.

\ignore{
\begin{acks}
We thank Nisarg Patel, ...
This work was partially funded by the National Science Foundation under the award ???.
\end{acks}
}

\clearpage
